---
title: "Brief Introduction to R Markdown"
author: "Joshua French"
date: "2023-01-18"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## What is R Markdown? 

R Markdown is a tool for dynamic, reproducible creation of data-centric documents that include text, code, the results of executed code.

- R Markdown was created by R Studio (<https://www.rstudio.com>) in 2014 (or at least that's the oldest discussion I can find).
- It is a variant of Markdown, which was created by John Gruber (creator of Daring Fireball (<https://daringfireball.net/>)) and Aaron Swartz (co-creator of Reddit ([https://www.reddit.com/](https://www.reddit.com)) as a lightweight markup language for creating formatted text using a plain-text editor.
- In general, a markdown language is a language that is written in plain-text but that is rendered different based on "markups" in the text.
- Has some capabilities similar to Jupyter (<https://jupyter.org/>) notebooks.

R Markdown is a great way to:

- Create a data analysis notebook that documents how and why you did analysis a certain way.
- Share your work with others in a variety of formats.
- Make your work reproducible.

R Markdown is a very useful tool for the data scientist.

R Studio provides R Markdown (and other) cheatsheets at <https://rstudio.com/resources/cheatsheets/>


## What's in an R Markdown document?

An R Markdown document has three components:

1.  An optional YAML header.
2.  Formatted text.
3.  **Chunks** of computer code.

YAML stands for 'Yet Another Markdown Language.'

## Formatting text in R Markdown

### Headings and sections

Headings and sections of different levels are created with `#`.

- `#` indicates the first level
- `##` being the second level
- etc.

The words after the `#` will be larger than the regular text.

### Font styles and effects

- *Italic text*: `*text*` or `_text_`.
- **Bold text**: `**text**` or `__text__` (two underscores on each side)
- `Typewriter (code) text`: `` `text` ``
- ^superscript^ and ~subscript~: `^superscript^` and `~subscript~`
- ~~strikethrough~~: `~~strikethrough~~`

### Numbered/ordered lists

Numbered lists with sub-bullets:

A numbered list with subnumbering.

1. Item 1
    a. Sub-item 1 # two tabs, not one
    b. Sub-item 2
        i. Sub-sub-item 1 # four tabs, not two
2. Item 2
    a. Sub-item 1

Note: There must be a line between any text and the number `1`.

### Unordered lists

Unordered lists with sub-bullets:

Some text with bullets.

* Item 1
  * Sub-item 1
  * Sub-item 2
    * Sub-sub-item 1
* Item 2
  * Sub-item 1

Notes:

- There must be a line between any text and the \*.
- You can use `+` or `-` instead of `*` when creating unordered lists with the same result.

### Code evaluation

- `eval`: (`TRUE`) A logical value indicating whether the chunk should be evaluated in the Console.
-  You can also evaluate **selected lines** if you provide a numeric vector. e.g., `eval = 1:3` would evaluate lines 1 through 3.

```{r, eval = TRUE}
seq(from = 3, to = 20, by = 4)
seq(3, 20, 4)
rnorm(n=10)
```

### Text output

- `echo`: (`TRUE`). A logical value indicating whether the source code should be displayed.
```{r, eval = TRUE, echo = FALSE}
seq(from = 3, to = 20, by = 4)
seq(3, 20, 4)
rnorm(n=10)
```
- You can also echo **selected lines** if you provide a numeric vector. e.g., `echo = 1:3` would echolines 1 through 3
- `results`: (`'markup'`). Determines how the results are displayed.
- Other options are `'asis'`, `'hold'`, `'hide'`.

```{r, eval = TRUE, echo = TRUE, results = 'asis'}
seq(from = 3, to = 20, by = 4)
seq(3, 20, 4)
rnorm(n=10)
```

- `collapse`: (`FALSE`). A logical value indicating whether the source and output should be collapsed into a single block.

```{r, eval = TRUE, echo = TRUE, results = 'markup', collapse = TRUE}
seq(from = 3, to = 20, by = 4)
seq(3, 20, 4)
rnorm(n=10)
```

- `include`: (`TRUE`) A logical value indicating whether to include the chunk output. Set to `FALSE` when you want something to run but not be shown.

```{r, eval = TRUE, echo = FALSE, results = 'asis', collapse = TRUE, include = FALSE}
seq(from = 3, to = 20, by = 4)
seq(3, 20, 4)
rnorm(n=10)
```

Other options: `warning`, `error`, `messages` (`TRUE`) determine whether warnings, errors, and messages will be displayed, respectively.

### Equations

Math is rendered using LaTeX (which is [pronounced](https://www.latex-project.org/about/) "Lah-tech" or "Lay-tech"). So you can use \$ and \$\$ to render inline math and display math like you would in LaTeX.

E.g., Typing "The regression coefficient \$\\beta\$" will produce "The regression coefficient $\beta$".

A "display style equation" is an equation that is set off from the rest of the text. The symbols are often larger, and the equation is often centered.

A display style equation is created between `\[ ... \]` or `$$ ... $$`. You add the various command to typeset your equation in the `...`. 

Consider adding the following text in your document:
$$
Y = \beta_0 + \beta_1 X + \epsilon.
$$
More details on how to type math on R Markdown can be found here <https://rpruim.github.io/s341/S19/from-class/MathinRmd.html>

This is my line and I show the same equation $Y = \beta_0$.

## YAML header basics 

The YAML (Yet Another Markdown Language) header controls many aspects of the R Markdown document.

We only discuss some of the basic details of YAML headers below. If you do need to know about the YAML header in detail, then you probably need a more advanced introduction to R Markdown!

The YAML header is provided at the top of an R Markdown document between sets of 3 dashes.

The YAML header allows you to specify:

- `title`: The title of the document, which is displayed in font larger than the section headings or main text.
- `author`: The author of the document. This is displayed just below the title.
- `date`: The date. You can use something like "2022-09-01" to add a static date.
- `output`: The output type produced when knitting the document.

### R Markdown output types 

One of the incredible features of R Markdown is that you can easily produce multiple output types from a single document. Some of the main document output types are:

- `pdf_document`: Excellent for distributing a static (unchanging) document. Also great for submitting homework solutions to your instructor! To compile into a pdf, you must install "tinytex."
```{r}
#install.packages("tinytex")
#tinytex::install_tinytex()
```

- `html_document`: Excellent for sharing interactive files. Can be viewed in a web browser.
- `word_document`: If for some reason you need Microsoft Word editing capabilities.

